Wait-free Queues with Polygarithmic Step Complexity

### 2022-07-13 article/psuedocode.tex
add detail dsearch

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}
\newcommand{\head}{head}
\newcommand{\size}{size }

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, i\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}
\State \tt{<b\sub{enq}, i\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
\label{deqRest}
\If{\tt{i\sub{enq}==-1}}
\State \tt{output= null}
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{Returns \tt{<b\sub{e},i\sub{e}>}, if $E_{root,b_e,i_e}$ is the response to the $D_{root,b_d,i_d}$. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - i\sub{d} $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\State \tt{r\sub{enq}= i\sub{d} - root.blocks[b\sub{d}-1].size + root.blocks[b\sub{d}-1].sum\sub{enq}}
\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
\EndIf
\EndFunction{FindResponse}

@@

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e} $\geq$ 1}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_{root,e}=E_{root,b,i}$.}
\State \tt{start= end-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[hbt]  
  \center\includegraphics[width=7in]{pics/blockSumEnq.png}

@@

\begin{lemma}[DSearch correctness] \label{dsearch}
If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E_{root, i}$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E_{root,b,i}$ in $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) steps.
\end{lemma}
\begin{proof}
First we show \nf{end-b}$\leq$\nf{root.blocks[b].\size+root.blocks[end].\size}. We know each block size is greater than 0. So every block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or one \nf{Dequeue()}. There cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the queue would become empty after bth block end before end and $E(n,i)$ could not be the response to to some \nf{DEQ} in \nf{end}. And since the lentgh of the queue would become \nf{root.blocks[end].\size} in the end so there cannot be more than \nf{root.blocks[end].\size} Dequeus in \nf{root.blocks[b..end]}. Cause if there was more then the end's lenght would become more than \nf{root.blocks[end].\size}.

Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} distance between end and b then with doubling search  in log\nf{root.blocks[b].\size+root.blocks[end].\size} steps we reach a block \nf{c} that the \nf{c.sum\sub{enq}} is less than i and the distance between c and end is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}. So the binary search takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps.
\end{proof}

@@

%\begin{lemma}[Search Ranges]\label{search}
%  Preconditions of all invocation of \nf{BSearch} are satisfied.
%\end{lemma}
%\begin{proof}
%  
%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
%
%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child. Search is done over the left subblocks. The start and end of the range are followed by definition. Line 92 is the same.
%
%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
%
%\end{proof}
We call the dequeues that return some value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= \size - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.
\begin{lemma}[Computing Queue's Head block] \label{computeHead}
  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it. \nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$. If the queue is empty in $S$ then it returns \nf{<-1,-->}.
  \end{lemma}
\begin{proof}
  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}. It is computed while creating the block in Line~\ref{computeLength}. If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0. Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size  of it.

>>>>

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\sc}[1]{\textsc{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}
\newcommand{\head}{head}
\newcommand{\size}{size }

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, i\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
\label{deqRest}
%\If{\tt{i\sub{enq}==-1}}
%\State \tt{output= null}
%\Else
%\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
%\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\Statex\cmt{Returns the the response to the $D_{root,b_d,i_d}$. Returns \tt{null} if the queue is empty.}
\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - i\sub{d} $<$ 0}}
\State \Return \tt{null}
\Else
\State \tt{r\sub{enq}= i\sub{d} - root.blocks[b\sub{d}-1].size + root.blocks[b\sub{d}-1].sum\sub{enq}}
\State \Return \tt{root.GetENQ(root.\Call{DSearch}{r\sub{enq}, b\sub{d}})}
\EndIf
\EndFunction{FindResponse}

@@

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_{root,e}=E_{root,b,i}$.}
\State \tt{start= end-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/blockSumEnq.png}

@@

\begin{lemma}[DSearch correctness] \label{dsearch}
Assume \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{e} and $E_{root, e}$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]}.  \nf{DSearch(e, end)} returns \nf{<b, i>} such $E_{root,b,i}= E_{root,e}$ in $\Theta \big(\log($\nf{root.blocks[b].\size + root.blocks[end].\size} \hspace{-0.8em}$)\big)$ steps.
\end{lemma}
\begin{proof}
First we show \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}. From line \ref{addOP}, we know that size of the every block in the tree is greater than 0. So each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or at least one \nf{Dequeue()}. There cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the  elements of the queue after \nf{blocks[b]} would become dequeued at some point, after \nf{root.blocks[b]} operations finish end before \nf{root.blocks[end]} operations start. And the response to to some \nf{dequeue()} in \nf{root.blocks[end]} could not be in $E(n,b)$. Furthermore since the size of the queue would become \nf{root.blocks[end].\size} in the end and there cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s so there cannot be more than \nf{root.blocks[b].\size + root.blocks[end].\size} \nf{enqueue()} operations in \nf{root.blocks[b..end]}. Cause if there were more, then the \nf{root.blocks[end]}'s size would become more than \nf{root.blocks[end].\size}. Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}. So \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}. See Figure \ref{end-b}.

Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}. Cause if otherwise, then \nf{(end-c)/2} satisfied the \nf{root.blocks[(end-c)/2].sum\sub{enq}<e}. Note that in the line \ref{doubling} the distance between \nf{end} and \nf{stat} is doubled. See figure \ref{fig::doubling}.

After computing \nf{b}, the value \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}). So the routine non constant part is the binary search which takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps from the first paragraph.
\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/doubling.png}

  \caption{Distance relations between $b,c,end$}
  \label{fig::doubling}
\end{figure}

\begin{figure}[hbt]  
  \center\includegraphics[width=3in]{pics/end-b.png}

  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
  \label{end-b}
\end{figure}

@@

%\begin{lemma}[Search Ranges]\label{search}
%  Preconditions of all invocation of \nf{BSearch} are satisfied.
%\end{lemma}
%\begin{proof}
%  
%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
%
%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child. Search is done over the left subblocks. The start and end of the range are followed by definition. Line 92 is the same.
%
%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
%
%\end{proof}
\it{Update the findresposne returning arguments.}

We call the dequeues that return some value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= \size - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.
\begin{lemma}[Computing Queue's Head block] \label{computeHead}
  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it. \nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$. If the queue is empty in $S$ then it returns \nf{<-1,-->}.
  \end{lemma}
\begin{proof}
  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}. It is computed while creating the block in Line~\ref{computeLength}. If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0. Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size  of it.
